Notes on Validated Model Counting \\ Version of \today

### 2022-04-27 doc/mc-notes.tex
Minor updates

{\bf Example 2}: Let $\ring$ be the set of rational numbers.  Assign a
{\em weight} $w_i$ to each variable $x_i$ such that $0 \leq w_i \leq
1$ and let $\cost(x_i) = w_i$.  This implements {\em weighted model counting}
under the restrictions that: 
1) the
weight of an assignment equals the product of the weights of its
literals, and 2) the weight of a variable $x_i$ and its complement
$\obar{x}_i$ sum to 1.

@@

\begin{proposition}[Complementation]
\label{prop:complementation}
  For separable cost function $\cost$ and Boolean function $f$:
  $\cost(\neg f) = 1 - \cost(f) = \oneminus \cost(f)$.
\end{proposition}

@@

We also find it useful to introduce a restricted version of the
if-then-else operator, denoted $\pite$.  For functions $f_1$,
$f_2$, and $f_3$, we define $\pite(f_1, f_2, f_3) = (f_1 \pand f_2)
\por (\neg f_1 \pand f_3)$.  The two arguments to the disjunction are
guaranteed to satisfy the requirements for an assignment-partitioned
disjunction, since the first argument can only be satisfied by assignments
for which $f_1$ evaluates to $\tautology$, while the second can only
be satisfied by assignments for which $f_1$ evaluates to $\nil$.  The
only restriction of the $\pite$ operator is that
$\depend(f_1) \cap \depend(f_2) = \emptyset$ and
$\depend(f_1) \cap \depend(f_3) = \emptyset$.

@@

\begin{table}
  \caption{Recursive Definition of Separable Schemas}
  \label{tab:schema}
A) Basic Operations\\
\begin{center}
\begin{tabular}{cccc}
\toprule
  \makebox[15mm]{$S$}           & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(S)$} & \makebox[30mm]{$\interpset{S}$}\\
\midrule
  $\nil$  & None  & $\emptyset$    & $\emptyset$ \\
  $\tautology$        & None  & $\emptyset$    & $\uinterp$ \\
$x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ \\
$\neg S_1$      & None  & $\depend(S_1)$ & $\uinterp - \interpset{S_1}$ \\
$S_1 \pand S_2$ & $\depend(S_1) \cap \depend(S_2) = \emptyset$ & $\depend(S_1) \cup \depend(S_2)$ & $\interpset{S_1} \cap \interpset{S_2}$ \\
$S_1 \por S_2$  & $\interpset{S_1} \cap \interpset{S_2} = \emptyset$ &  $\depend(S_1) \cup \depend(S_2)$ &  $\interpset{S_1} \cup \interpset{S_2}$ \\
\bottomrule
\end{tabular}
\end{center}
B) Composite Operations\\
\begin{center}
\begin{tabular}{cc}
\toprule
  $S$           & Definition \\
\midrule
\makebox[30mm]{$\pite(S_1, S_2, S_3)$} & \makebox[40mm]{$[S_1 \pand S_2] \por [\neg S_1 \pand S_2]$}\\
\bottomrule
\end{tabular}
\end{center}
\end{table}

A {\em separable schema} is a directed acyclic graph representing a
Boolean formula where the only allowed operations are $\neg$, $\pand$,
$\por$ and $\pite$.  We use a graph representation to allow sharing of
subformulas.  The set of schemas over a set of variables
$\{x_1, x_2, \ldots, x_n\}$
can be defined recursively, as is shown in Table
\ref{tab:schema}.  Each schema $S$ has an associated dependency set
$\depend(S)$ and an associated set of models $\interpset{S}$.  We
divide the operations into {\em basic} ones that are fundamental to
the representation, and {\em composite} ones that can be constructed
from basic operations.  The latter category consists of just the $\pite$ operation.

@@

\begin{table}
  \caption{Normalization Rules}
  \label{tab:normalize}
  \begin{center}
  \begin{tabular}{ccccccc}
    \makebox[20mm]{$\neg \nil$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\tautology$} & \makebox[15mm]{} &
    \makebox[20mm]{$\neg \tautology$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\nil$} \\
    $\neg \neg S$ & $\rightarrow$ & $S$ \\
    $S \pand \nil$ & $\rightarrow$ & $\nil$ && $\nil \pand S$ & $\rightarrow$ & $\nil$ \\
    $S \pand \tautology$ & $\rightarrow$ & $S$ && $\tautology \pand S$ & $\rightarrow$ & $S$ \\
    $S \por \nil$ & $\rightarrow$ & $S$ && $\nil \por S$ & $\rightarrow$ & $S$ \\
    $S \por \tautology$ & $\rightarrow$ & $\tautology$ && $\tautology \por S$ & $\rightarrow$ & $\tautology$ \\
    $\pite(\neg S_1, S_2, S_3)$ & $\rightarrow$ & $\pite(S_1, S_3, S_2)$ \\
    $\pite(\tautology, S_2, S_3)$ & $\rightarrow$ & $S_2$ & &
    $\pite(\nil, S_2, S_3)$ & $\rightarrow$ & $S_3$ \\
    $\pite(S_1, \tautology, \nil)$ & $\rightarrow$ & $S_1$ &&
    $\pite(S_1, \nil, \tautology)$ & $\rightarrow$ & $\neg S_1$ \\
    $\pite(S_1, S_2, \nil)$ & $\rightarrow$ & $S_1 \pand S_2$ & &
    $\pite(S_1, \nil, S_3)$ & $\rightarrow$ & $\neg S_1 \pand S_3$ \\
    $\pite(S_1, S_2, \tautology)$ & $\rightarrow$ & $\neg [S_1 \pand \neg S_2]$ & &
    $\pite(S_1, \tautology, S_3)$ & $\rightarrow$ & $\neg [\neg S_1 \pand \neg S_3]$ \\
  \end{tabular}
  \end{center}
\end{table}

Table~\ref{tab:normalize} shows a list of {\em normalizing}
transformations to simplify a separable schema.  These are mostly
straightforward---eliminating extra negations and removing constant
terms.  The most interesting are the two bottom rules, which
use DeMorgan's Laws to simplify an $\pite$ operation when one of
the arguments is $\tautology$ into a $\pand$ operation.  We will make use of this
transformation when using BDDs to convert a set of clauses into a
separable schema.

@@

The CRAT clausal proof framework provides a means for creating a
checkable proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to some separable schema.  This provides
a framework for adding proof generation to model counting programs.

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
    $C$ & {\tt ab} & $L^{+}$ {\tt 0} & ${\tt -}C^{*}$  {\tt 0}  & Add blocked clause \\
    $C$ & {\tt ar} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RAT clause \\
     & {\tt db} & $C$             & ${\tt -}C^{+}$  {\tt 0} & Delete blocked clause \\
     & {\tt dr} & $C$             & $C^{+}$  {\tt 0} & Delete RAT clause \\
    \midrule
        & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
        & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
  \end{tabular}
  }
\end{table}

@@

The first set of proof rules are similar to those in other clausal
proofs.  Our syntax optionally allows input clauses to be listed with
a rule of type {\tt i}.  Clauses can be added via a blocked-clause addition (command {\tt ab})
or a RAT addition (command {\tt dr}).
As described below, however, blocked clauses can only be added
to define $\pand$, $\por$, and $\pite$ operations.  The hints portion
of a blocked-clause addition lists all earlier clauses containing the
negated version of the pivot literal, with the clause IDs negated.
The hints portion of a RAT addition must contain a sequence of clause
IDs such that the added clause is RAT with respect to these clauses.
Similarly, a clause can be deleted if it blocked with respect to one
of its literals (command {\tt db}), or because it is RAT (command {\tt
  dr}) The hints portion of a RAT deletion operation must be an
ordered list of clause IDs justifying the deletion.

@@

An if-then-else rule of the form ${\tt ite}\;v\;l_1\;l_2\;l_3$
indicates that $v$ will represent $\pite(l_1, l_2, l_3)$.  The blocked
clause additions must encode the formula
$v \leftrightarrow \ite((l_1,l_2, l_3)$.
The dependency set for $l_1$ must be disjoint from those
of $l_2$ and $l_3$.

@@

A CRAT proof follows the same general form as a QRAT dual proof---one
that ensures that each clause addition and each clause deletion
preserves equivalence.  Starting with the set of input clauses, it
produces a sequence of steps that both add and delete clauses.  Each
addition must be truth preserving and each deletion must be falsehood
preserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause
consisting of some variable or its complement.  Except for trivial
cases, the final literal will be an extension variable or its
complement.  That literal will indicate the root of the schema as the
(possibly negated) outpuf of a schema operation.  Working from that
root backward, the schema can be extracted from the CRAT file.

@@

As a more complex example, consider the Boolean formula given by the
conjunction of clauses $C_1 = x_1 \lor x_2 \lor x_3$ and $C_2 =
\obar{x}_1 \lor x_2$.  With this example, we also demonstrate the use
of DeMorgan's Laws to provide a more compact encoding of the formula.
The complement of the formula can be written in DNF as
$\obar{x}_1\,\obar{x}_2\,\obar{x}_3 \lor x_1\,\obar{x}_2$.  Each of
the two conjuncts can be formed using $\pand$ operations, and their
sum can be formed using a $\por$ operation.  In this example, we first
form a term $t_1$ that equals $C_1$ and a term $t_2$ that equals
$C_2$.  These are asserted as unit clauses in proof lines 9 and 13,
allowing the input clauses to be deleted.  Then we generate $t_3$ as
the conjunction of $t_1$ and $t_2$ and assert it as a unit clause.
Based on this, we can delete the unit clauses for terms $t_1$ and
$t_2$.  Term $t_3$ then becomes the root of the schematic representation.

@@

Given an arbitrary CNF formula, we can use BDD operations to generate
a schematic representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.  The key idea is to use
extended resolution to encode the semantics of the BDD nodes as part
of the proof.  Each BDD node defines an ITE operation.  As the
transformations of Table~\ref{tab:normalize} indicate, each of these
will be represented in the proof by a literal, an $\pand$ operation,
or an $\pite$ operation.  The ITE operation for a node has an input
variable $x_i$ as its first argument.  In an ordered BDD, the
dependency sets of the two children will include only variables
greater than $x_i$ in the ordering.  Therefore, these operations will
satisfy the partitioned-variable requirements of the  $\pand$ and $\pite$ operations.

Other model counters use either top-down and bottom-up methods to
generate representations of the formula that are similar to our
schmas.  These exploit the key abstractions we have identified.  We
must find ways to modify these model counters to also generate CRAT
proofs.

>>>>

{\bf Example 2}: Let $\ring$ be the set of rational numbers.  Assign a
{\em weight} $w_i$ to each variable $x_i$ such that $0 \leq w_i \leq
1$ and let $\cost(x_i) = w_i$.  This implements {\em weighted model counting}
under the restrictions that: 
1) the
weight of an assignment equals the product of the weights of its
literals, and 2) the weight of a variable $x_i$ and its negation
$\obar{x}_i$ sum to 1.

@@

\begin{proposition}[Negation]
\label{prop:negation}
  For separable cost function $\cost$ and Boolean function $f$:
  $\cost(\neg f) = 1 - \cost(f) = \oneminus \cost(f)$.
\end{proposition}

@@

%% We also find it useful to introduce a restricted version of the
%% if-then-else operator, denoted $\pite$.  For functions $f_1$,
%% $f_2$, and $f_3$, we define $\pite(f_1, f_2, f_3) = (f_1 \pand f_2)
%% \por (\neg f_1 \pand f_3)$.  The two arguments to the disjunction are
%% guaranteed to satisfy the requirements for an assignment-partitioned
%% disjunction, since the first argument can only be satisfied by assignments
%% for which $f_1$ evaluates to $\tautology$, while the second can only
%% be satisfied by assignments for which $f_1$ evaluates to $\nil$.  The
%% only restriction of the $\pite$ operator is that
%% $\depend(f_1) \cap \depend(f_2) = \emptyset$ and
%% $\depend(f_1) \cap \depend(f_3) = \emptyset$.

@@

\begin{table}
  \caption{Recursive Definition of Separable Schemas}
  \label{tab:schema}
%%A) Basic Operations\\
\begin{center}
\begin{tabular}{cccc}
\toprule
  \makebox[15mm]{$S$}           & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(S)$} & \makebox[30mm]{$\interpset{S}$}\\
\midrule
  $\nil$  & None  & $\emptyset$    & $\emptyset$ \\
  $\tautology$        & None  & $\emptyset$    & $\uinterp$ \\
$x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ \\
$\neg S_1$      & None  & $\depend(S_1)$ & $\uinterp - \interpset{S_1}$ \\
$S_1 \pand S_2$ & $\depend(S_1) \cap \depend(S_2) = \emptyset$ & $\depend(S_1) \cup \depend(S_2)$ & $\interpset{S_1} \cap \interpset{S_2}$ \\
$S_1 \por S_2$  & $\interpset{S_1} \cap \interpset{S_2} = \emptyset$ &  $\depend(S_1) \cup \depend(S_2)$ &  $\interpset{S_1} \cup \interpset{S_2}$ \\
\bottomrule
\end{tabular}
\end{center}
%% B) Composite Operations\\
%% \begin{center}
%% \begin{tabular}{cc}
%% \toprule
%%   $S$           & Definition \\
%% \midrule
%% \makebox[30mm]{$\pite(S_1, S_2, S_3)$} & \makebox[40mm]{$[S_1 \pand S_2] \por [\neg S_1 \pand S_2]$}\\
%% \bottomrule
%% \end{tabular}
%% \end{center}
\end{table}

A {\em separable schema} is a directed acyclic graph representing a
Boolean formula where the only allowed operations are $\neg$, $\pand$,
$\por$.
%% and $\pite$.
We use a graph representation to allow sharing of
subformulas.  The set of schemas over a set of variables
$\{x_1, x_2, \ldots, x_n\}$
can be defined recursively, as is shown in Table
\ref{tab:schema}.  Each schema $S$ has an associated dependency set
$\depend(S)$ and an associated set of models $\interpset{S}$.
%% We
%% divide the operations into {\em basic} ones that are fundamental to
%% the representation, and {\em composite} ones that can be constructed
%% from basic operations.  The latter category consists of just the $\pite$ operation.

@@

\begin{table}
  \caption{Normalization Rules}
  \label{tab:normalize}
  \begin{center}
  \begin{tabular}{ccccccc}
    \makebox[20mm]{$\neg \nil$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\tautology$} & \makebox[15mm]{} &
    \makebox[20mm]{$\neg \tautology$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\nil$} \\
    $\neg \neg S$ & $\rightarrow$ & $S$ \\
    $S \pand \nil$ & $\rightarrow$ & $\nil$ && $\nil \pand S$ & $\rightarrow$ & $\nil$ \\
    $S \pand \tautology$ & $\rightarrow$ & $S$ && $\tautology \pand S$ & $\rightarrow$ & $S$ \\
    $S \por \nil$ & $\rightarrow$ & $S$ && $\nil \por S$ & $\rightarrow$ & $S$ \\
    $S \por \tautology$ & $\rightarrow$ & $\tautology$ && $\tautology \por S$ & $\rightarrow$ & $\tautology$ \\
%%     $\pite(\neg S_1, S_2, S_3)$ & $\rightarrow$ & $\pite(S_1, S_3, S_2)$ \\
%%     $\pite(\tautology, S_2, S_3)$ & $\rightarrow$ & $S_2$ & &
%%     $\pite(\nil, S_2, S_3)$ & $\rightarrow$ & $S_3$ \\
%%     $\pite(S_1, \tautology, \nil)$ & $\rightarrow$ & $S_1$ &&
%%     $\pite(S_1, \nil, \tautology)$ & $\rightarrow$ & $\neg S_1$ \\
%%     $\pite(S_1, S_2, \nil)$ & $\rightarrow$ & $S_1 \pand S_2$ & &
%%     $\pite(S_1, \nil, S_3)$ & $\rightarrow$ & $\neg S_1 \pand S_3$ \\
%%     $\pite(S_1, S_2, \tautology)$ & $\rightarrow$ & $\neg [S_1 \pand \neg S_2]$ & &
%%     $\pite(S_1, \tautology, S_3)$ & $\rightarrow$ & $\neg [\neg S_1 \pand \neg S_3]$ \\
  \end{tabular}
  \end{center}
\end{table}

Table~\ref{tab:normalize} shows a list of {\em normalizing}
transformations to simplify a separable schema.  These 
eliminate extra negations and remove constant
terms, such that constants only occur in schemas when representing constant functions $\zero$ and $\one$.
%% The most interesting are the two bottom rules, which
%% use DeMorgan's Laws to simplify an $\pite$ operation when one of
%% the arguments is $\tautology$ into a $\pand$ operation.  We will make use of this
%% transformation when using BDDs to convert a set of clauses into a
%% separable schema.

@@

The CRAT clausal proof framework provides a means for creating a
checkable proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to a separable schema.  Once this
equivalence has been established, the schema can form the basis
for computations enabled by the
this representation, including trusted model counting

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
    $C$ & {\tt ab} & $L^{+}$ {\tt 0} & ${\tt -}C^{*}$  {\tt 0}  & Add blocked clause \\
    $C$ & {\tt ar} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RAT clause \\
     & {\tt db} & $C$             & ${\tt -}C^{+}$  {\tt 0} & Delete blocked clause \\
     & {\tt dr} & $C$             & $C^{+}$  {\tt 0} & Delete RAT clause \\
    \midrule
        & {\tt p} & $V \; L \; L$    &                  & Declare $\pand$ operation \\
        & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation \\ 
%%        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
  \end{tabular}
  }
\end{table}

@@

The first set of proof rules are similar to those in other clausal
proofs.  Our syntax optionally allows input clauses to be listed with
a rule of type {\tt i}.  Clauses can be added via a blocked-clause addition (command {\tt ab})
or a RAT addition (command {\tt ar}).
As described below, however, blocked clauses can only be added
to define $\pand$ and $\por$
%%, and $\pite$
operations.  The hints portion
of a blocked-clause addition lists all earlier clauses containing the
negated version of the pivot literal, with the clause IDs negated.
The hints portion of a RAT addition must contain a sequence of clause
IDs such that the added clause is RAT with respect to these clauses.
Similarly, a clause can be deleted if it blocked with respect to one
of its literals (command {\tt db}), or because it is RAT (command {\tt
  dr}) The hints portion of a RAT deletion operation must be an
ordered list of clause IDs justifying the deletion.

@@

%% An if-then-else rule of the form ${\tt ite}\;v\;l_1\;l_2\;l_3$
%% indicates that $v$ will represent $\pite(l_1, l_2, l_3)$.  The blocked
%% clause additions must encode the formula
%% $v \leftrightarrow \ite((l_1,l_2, l_3)$.
%% The dependency set for $l_1$ must be disjoint from those
%% of $l_2$ and $l_3$.

@@

A CRAT proof follows the same general form as a QRAT dual proof~\cite{bryant:cade:2021}---one
that ensures that each clause addition and each clause deletion
preserves equivalence.  Starting with the set of input clauses, it
produces a sequence of steps that both add and delete clauses.  Each
addition must be truth preserving and each deletion must be falsehood
preserving.  At the end, all input clauses must have been deleted, and
among the remaining clauses there must be only a single unit clause
consisting of some variable or its complement.  Except for trivial
cases, the final literal will be an extension variable or its
complement.  That literal will indicate the root of the schema as the
(possibly negated) outpuf of a schema operation.  Working from that
root backward, the schema can be extracted from the CRAT file.

@@

As a more complex example, consider the Boolean formula given by the
conjunction of clauses $C_1 = x_1 \lor x_2 \lor x_3$ and $C_2 =
\obar{x}_1 \lor x_2$.  With this example, we also demonstrate the use
of DeMorgan's Laws to provide a more compact encoding of the formula.

The complement of the formula can be written in DNF as
$\obar{x}_1\,\obar{x}_2\,\obar{x}_3 \lor x_1\,\obar{x}_2$.  Each of
the two conjuncts can be formed using $\pand$ operations, and their
sum can be formed using a $\por$ operation.  In this example, we first
form a term $t_1$ that equals $C_1$ and a term $t_2$ that equals
$C_2$.  (Note that $t_1$ in this case is logically equivalent to $t_1$
in the example of Figure~\ref{fig:c3:schema}, but the use of
negation enables it to be represented with two operations
rather than four.)
Terms $t_1$ and $t_2$ are asserted as unit clauses in proof lines 9 and 13,
allowing the input clauses to be deleted.  Then we generate $t_3$ as
the conjunction of $t_1$ and $t_2$ and assert it as a unit clause.
Based on this, we can delete the unit clauses for terms $t_1$ and
$t_2$.  Term $t_3$ then becomes the root of the schematic representation shown in Figure~\ref{fig:p2:schema}.

@@

Given an arbitrary CNF formula, we can use BDD operations to generate
a schematic representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.  The key idea is to use
extended resolution to encode the semantics of the BDD nodes as part
of the proof.  Depending on which, if any, of its children are
terminal nodes, a BDD node will be represented in the schema as either
1) a literal, 2) a single $\pand$ operation, or 3) two $\pand$
operations and one $\por$ operation.

Other model counters use either top-down and bottom-up methods to
generate representations of the formula that are similar to our
schemas.  These exploit the key abstractions we have identified.  We
must find ways to modify these model counters to also generate CRAT
proofs.
